---
title: "Lecture 9: Computation with Random Variables"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
```


## R preparation

### Defining custom functions

In R, you can define your own function to perform a set of tasks. You can
define the set of inputs (_arguments_) and its output (_return value_). After defining
your own function, you can call function to peform the specified set of
tasks repeatedly.

```{r eval=FALSE}
my_function <- function(arg1, arg2, arg3 = < default value >) {
  < define task(s) >
  return(< output value >)
}
```

+  `my_function <- function()` assigns the defined function to variable `my_function`. 
You can name a function just like any variable - start with a letter and can be
a combination of letters, digits, period (`.`), and underscore (`_`) without any spaces.  
I suggest following a well-established convention such
as [snake_case](https://en.wikipedia.org/wiki/Snake_case) in a consistent manner
and giving a brief but descriptive name.

+  `function(arg1, arg1, arg3 = < default value >)` describes the set of input
arguments the function takes. Arguments without "=" assignments are mandatory
input arguments and users are expected to provide values for all of them.
Arguments defined with "=" assignments are optional input arguments that will
take the assigned value as the default value if the user doesn't provide any
value for the arguments. Optional arguments must be defined after the mandatory
arguments.

+  `return(< output value>)` defines the return value of the function. When you
call the function, it returns the value inside `return()` or if there is no
explicit `return()`, the result of the last statement. While you can define
the return value without explicitly calling `return()`, it can make the function
easier to read and understand.

For example, I can define a function that prints and returns the information
provided as input arguments.

```{r echo=TRUE}
record_info <- function(name, sex, is_student = TRUE) {
  print(paste(name, "is a", sex, 
              ifelse(is_student, "student.", "and not a student.")))
  record <- data.frame(
    name = name, sex = sex, is_student = is_student
  )
  return(record)
}
```

```{r record-setup}
record_info <- function(name, sex, is_student = TRUE) {
  print(paste(name, "is a", sex, 
              ifelse(is_student, "student.", "and not a student.")))
  record <- data.frame(
    name = name, sex = sex, is_student = is_student
  )
  return(record)
}
```

The function `record_info()` is available in the code chunk below. Test the function
with your information.

```{r record, exercise=TRUE}
record_info("Michael", "male", FALSE)
```


## Blackjack against a dealer

The game of Blackjack is a game played with one more decks of playing cards.
Below is a single deck of playing cards.

```{r echo=TRUE}
suits <- c("Spades", "Hearts", "Clubs", "Diamonds")
numbers <- c("A", 2:10, "J", "Q", "K")
values <- c(11, 2:10, 10, 10, 10)
deck <- data.frame(
  suit = rep(suits, each = 13),
  number = rep(numbers, 4),
  value = rep(values, 4)
)
deck
```

In a game of Blackjack, one ore more players play against a dealer. Scores for the player
and the dealer are counted using the rule described below.

_Counting rule:_

+  Scores are based on the `number`s.
+  Numbers between `2` and `10` are always counted as is.
+  Count the face cards where the `number` is `J`, `Q`, or `K` as $10$.
+  Count $A$s on hand as $11$ unless the total score exceeds $21$.
    a.  When the total score exceeds $21$ and there is an $A$ on hand counted as $11$, count the $A$ as $1$.
    b.  Repeat the above step until the total score is less than or equal to $21$ or there are no more $A$ counted as $11$ on hand.


_Winning objective:_

+  The objective of the game is to score higher than the dealer without exceeding $21$.
    a. When a player scores $21$ with their first pair, the player wins immediately.
    b. When a player exceeds $21$ after counting all $A$s on hand as $1$, the player loses immediately.
+  For simplicity, we will consider all tied scores a win for the player even when the dealer scores $21$ with their first pair.

_Game play:_

+  The game starts with each player and the dealer getting 2 pairs of cards
drawn without replacements from a randomly shuffled deck(s) of playing cards.
The cards of each player is revealed to everyone while only one of the dealer's
card is revealed to the players.

```{r echo=FALSE}
set.seed(237)
shuffled <- sample(seq_len(52))
player <- 1:2
dealer <- 3:4
blackjack_table <- ggplot() +
  theme_void() +
  scale_x_continuous(limits = c(0, 3)) +
  scale_y_reverse(limits = c(8, -.5)) +
  annotate("rect", xmin = 0, xmax = 3, ymin = 8, ymax = -.5, fill = NA, color = "darkgrey") +
  annotate("text", x = 1:2, y = 0, label = c("Player", "Dealer"), size = 5, vjust = 1, hjust = .5) +
  annotate("segment", x = 1.5, xend = 1.5, y = 8, yend = -.5, linetype = "dotted")
blackjack_table +
  annotate("text", x = 1, y = player, 
           label = paste(deck$number[shuffled[player]], deck$suit[shuffled[player]]),
           color = ifelse(deck$suit[shuffled[player]] %in% c("Hearts", "Diamonds"), "red", "black")) +
  annotate("text", x = 2, y = dealer, 
           label = c(paste(deck$number[shuffled[dealer[1]]], deck$suit[shuffled[dealer[1]]]), "?"),
           color = c(ifelse(deck$suit[shuffled[dealer[1]]] %in% c("Hearts", "Diamonds"), "red", "black"), "black"))
```


+  Each player then decides whether to receive more cards one at a time. For example, the player may decide to draw one more card and then another.

```{r echo=FALSE}
player <- c(1:2, 5:6)
dealer <- 3:4
blackjack_table +
  annotate("text", x = 1, y = player, 
           label = paste(deck$number[shuffled[player]], deck$suit[shuffled[player]]),
           color = ifelse(deck$suit[shuffled[player]] %in% c("Hearts", "Diamonds"), "red", "black")) +
  annotate("text", x = 2, y = dealer, 
           label = c(paste(deck$number[shuffled[dealer[1]]], deck$suit[shuffled[dealer[1]]]), "?"),
           color = c(ifelse(deck$suit[shuffled[dealer[1]]] %in% c("Hearts", "Diamonds"), "red", "black"), "black"))
```

+  The player stops and scores $8+2+3+1=14$. If there is another player, the player makes their decisions. When all players finish making their decisions, the dealer reveals their second card.


```{r echo=FALSE}
player <- c(1:2, 5:6)
dealer <- 3:4
blackjack_table +
  annotate("text", x = 1, y = player, 
           label = paste(deck$number[shuffled[player]], deck$suit[shuffled[player]]),
           color = ifelse(deck$suit[shuffled[player]] %in% c("Hearts", "Diamonds"), "red", "black")) +
  annotate("text", x = 2, y = dealer, 
           label = paste(deck$number[shuffled[dealer]], deck$suit[shuffled[dealer]]),
           color = ifelse(deck$suit[shuffled[dealer]] %in% c("Hearts", "Diamonds"), "red", "black"))
```

+  The dealer draws cards until they score $17$ or higher.

```{r echo=FALSE}
player <- c(1:2, 5:6)
dealer <- c(3:4, 7)
blackjack_table +
  annotate("text", x = 1, y = player, 
           label = paste(deck$number[shuffled[player]], deck$suit[shuffled[player]]),
           color = ifelse(deck$suit[shuffled[player]] %in% c("Hearts", "Diamonds"), "red", "black")) +
  annotate("text", x = 2, y = dealer, 
           label = paste(deck$number[shuffled[dealer]], deck$suit[shuffled[dealer]]),
           color = ifelse(deck$suit[shuffled[dealer]] %in% c("Hearts", "Diamonds"), "red", "black"))
```

+  If the dealer's score exceeds $21$, all players win. In this example, the dealer scores $3+4+10=17$ which does not exceed $21$ and is higher than the player's score of $14$. The player loses.


### Defining the scoring function

The function `score_blackjack()` takes a subset of the `deck` data frame as
an input to score the hand.

```{r}
score_blackjack <- function(cards) {
  s <- sum(cards$value)
  n_aces <- sum(cards$value == 11)
  s <- s - 10 * min(ceiling((s - 21) / 10), n_aces)
  return(s)
}
```


### Defining the dealer

Since we know the rules for the dealer, we can implement how the dealer
decides to draw more cards. Write the function such that it returns `TRUE` 
if the decision is to draw one more card given the `cards` on the dealer's 
hand and `FALSE` otherwise.

Use `score_blackjack()` to write the function and data frame `deck` to test it. 
They are available in the code chunk below.

```{r define-dealer-setup}
suits <- c("Spades", "Hearts", "Clubs", "Diamonds")
numbers <- c("A", 2:10, "J", "Q", "K")
values <- c(11, 2:10, 10, 10, 10)
deck <- data.frame(
  suit = rep(suits, each = 13),
  number = rep(numbers, 4),
  value = rep(values, 4)
)
score_blackjack <- function(cards) {
  s <- sum(cards$value)
  n_aces <- sum(cards$value == 11)
  s <- s - 10 * min(ceiling((s - 21) / 10), n_aces)
  return(s)
}
```

```{r define-dealer, exercise=TRUE, exercise.lines=10}
dealer <- function(cards) {
  
}
```

```{r define-dealer-solution}
dealer <- function(cards) {
  decision <- score_blackjack(cards) < 17
  return(decision)
}
```


### Playing a game of Blackjack

In the code chunk below, the functions `score_blackjack()`, `dealer()`, 
`winner()`, and `player()` as well as the data frame `deck` are available. 

The function `winner()` takes the player's current score and the dealer's current
score as the input arguments and returns `TRUE` if the player wins and `FALSE` 
otherwise. The function `player()` expects the player's current hand as a subset
of `deck` and returns `TRUE` if the current hand's score is less than 20.

Let's write a game of Blackjack using the functions and indicate
whether the player won.

```{r play-blackjack-setup}
suits <- c("Spades", "Hearts", "Clubs", "Diamonds")
numbers <- c("A", 2:10, "J", "Q", "K")
values <- c(11, 2:10, 10, 10, 10)
deck <- data.frame(
  suit = rep(suits, each = 13),
  number = rep(numbers, 4),
  value = rep(values, 4)
)
score_blackjack <- function(cards) {
  s <- sum(cards$value)
  n_aces <- sum(cards$value == 11)
  s <- s - 10 * min(ceiling(max(0, (s - 21) / 10)), n_aces)
  return(s)
}
dealer <- function(cards) {
  decision <- score_blackjack(cards) < 17
  return(decision)
}
player <- function(cards){
  decision <- score_blackjack(cards) < 20
  return(decision)
}
winner <- function(player_score, dealer_score) {
  player_won <- TRUE
  if (dealer_score < 22 && dealer_score > player_score) {
    player_won <- FALSE
  }
  if (player_score > 21) {
    player_won <- FALSE
  }
  return(player_won)
}
```

```{r play-blackjack, exercise=TRUE, exercise.lines=15}
shuffled <- deck[sample(seq_len(nrow(deck))), ]
player_hand <- #
dealer_hand <- #
```

```{r play-blackjack-hint-1}
shuffled <- deck[sample(seq_len(nrow(deck))), ]
# deal first pairs
player_hand <- c(1:2)
dealer_hand <- c(3:4)
position <- 5 # track the top of the undealt deck
while (player(shuffled[player_hand, ])) {
  player_hand <- c(player_hand, position)
  position <- position + 1
}
# if you are writing a game, you should stop when the player's hand
# is either exactly 21 or exceeds 21
while (dealer(shuffled[dealer_hand, ])) {
  dealer_hand <- c(dealer_hand, position)
  position <- position + 1
}
player_score <- score_blackjack(shuffled[player_hand, ])
dealer_score <- score_blackjack(shuffled[dealer_hand, ])

shuffled[player_hand, ]
shuffled[dealer_hand, ]
                                
print(paste(
  ifelse(winner(player_score, dealer_score), "The player won,", "The player lost,"),
  player_score, "vs.", dealer_score))
```

```{r play-blackjack-hint-2}
shuffled <- deck[sample(seq_len(nrow(deck))), ]
play_a_round <- function(shuffled, position = 1) {
  # takes shuffled deck and the current position
  # of the top of the undealt deck as input arguments
  # deal first pairs
  player_hand <- c(position, position + 1)
  dealer_hand <- c(position + 2:3)
  position <- position + 4
  while (player(shuffled[player_hand, ])) {
    player_hand <- c(player_hand, position)
    position <- position + 1
  }
  player_score <- score_blackjack(shuffled[player_hand, ])
  if (player_score == 21) {
    return(c(TRUE, position)) # player wins immediately
  }
  if (player_score > 21) {
    return(c(FALSE, position)) # player loses immediately
  }
  while (dealer(shuffled[dealer_hand, ])) {
    dealer_hand <- c(dealer_hand, position)
    position <- position + 1
  }
  dealer_score <- score_blackjack(shuffled[dealer_hand, ])
  # return c(winner, current position)
  return(c(winner(player_score, dealer_score), position))
}
                                
print(paste(
  ifelse(play_a_round(shuffled)[1], "The player won.", "The player lost.")))
```
