---
title: "Lecture 9: Computation with Random Variables"
output: 
  learnr::tutorial:
    css: css/styles.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
```


## Example: Bike tires

Recall the bike tires example. $X$ is the right tire's pressure and $Y$ is the left tire's pressure in psi. They need to be filled to a pressure of 40 psi. You pick a tire randomly, represented by $W=0$ for picking the right tire and $W=1$ for picking the left tire. $V$ is the pressure measurement of the randomly selected tire. $(X,Y)$ has the following joint probability density function $f$.

$$f(x,y)=\begin{cases}K(x^2+y^2) & 20\le x,y\le50 \\ 0 &\text{otherwise}\end{cases}$$

where $K=2\ 340\ 000^{-1}$ and $W\sim\text{Ber}(1/2)$.

Recall that

$$ f_X(x)=\begin{cases}\frac{x^2}{78\ 000} +\frac{13}{780} & 20\le x \le 50 \\ 0 &\text{otherwise.}\end{cases}\quad$$

and

$$ F_X(a)=\begin{cases}0 & x <20 \\\frac{1}{234\ 000}a^3+\frac{1}{60}a-\frac{43}{117} & 20\le a \le 50 \\ 1 & x > 50\end{cases}$$

1. Simulate 100,000 samples of $V$
2. Plot a histogram of the samples
3. Estimate $E(V)$ and $P(V<40)$

```{r pv40, exercise=TRUE, exercise.lines=30}
N <- 100000
U <- runif(N)
cdfX <- function(a) {
  ifelse(
    a <= 20, 0, ifelse(
      a > 50, 1, 
      (a^3 + 3900 * a - 43 *2000)/ 234000
    )
  )
}
X <- numeric(N)

```



## Computation with normal random variables


For the following exercises, we use $X_1$, $X_2$, $X_3$, ..., $X_{25}$, where

$$X_i\sim N(0,1)$$

for all $i$ and they are independent.


### Simulating from a Single $X_1$

1. Simulate 100 samples of $X_1$ and 
    1)  plot their histogram;
    2)  compute their mean; and
    3)  compute their variance.

```{r ex1, exercise=TRUE, exercise.lines=5}
N <- 100
```

2. Simulate 100 samples of $3X_1 + 2$ and
    1)  plot their histogram;
    2)  compute their mean; and
    3)  compute their variance.

How do they compare to the results from Part 1.?

```{r ex2, exercise=TRUE, exercise.lines=5}
N <- 100
```

3. Simulate 100 samples of $\overline{X}$, where

$$\overline{X}=\frac{\sum_{i=1}^{25}}{25}$$

and 
    1)  plot their histogram;
    2)  compute their mean; and
    3)  compute their variance.

How do they compare to the results from Part 1.?

```{r ex3, exercise=TRUE, exercise.lines=10}
N <- 100
```

